Predicting the effects of character displacement on food-web dynamics

### 2019-03-07 manuscript_ECD_model.Rmd
delete old text I was keeping as notes

<!-- Outline

Introduction

1. ECD is an important process, and current empirical and theoretical work suggests it often results in divergence of consumer foraging traits.

2. Ecological consequences of character displacement have received surprisingly little attention. In particular, prior work doesn't allow a proper comparision of the effects of character displacement on community stability.

-> Another paragraph illustrating how the direct and indirect effects of character displacement likely have constrasting effects on stability.

3. Here we fill this knowledge gap. We focus on the ecological side of these dynamics, because the evolutionary aspects have been well characterized, although we show in the appendices that the evolutionary dynamics are as expected (especially for McCann model).

Methods and Results

- Underlying model of consumer-resource dynamics
  1. Explanation of model parameters
  2. Methods for assessing resource abundances and stability.
  3. Trade-off function and definition of character displacement.
  4. Analytical and simulation methods. Briefly mention the different scenarios I'll explore.

- Consumers can forage for resources simultaneously
  1. Functional response equation and illustration of assumptions
  2. Effect on attack rates, resource densities, and stability

- Consumers cannot foraging for resource simultaneously
  1. Functional response equation and illustration of assumptions
  2. Effect on attack rates, resource densities, and stability

- McCann's most realistic functional response
  1. Functional response equation and illustration of assumptions
  2. Effect on attack rates, resource densities, and stability

Discussion

-->

@@

Ecological character displacement is an adaptive process, often enhancing coexistence between consumers by reducing competition for shared resources. 
This focus on competition has undermined the fact that consumer-resource interactions ultimately mediate the process of character displacement. 
As a consequence, the effects of character displacement on consumer-resource interactions and food-web dynamics have not been critically examined.
Here, I study a model of two consumers competing for two shared resources to examine how character displacement in consumers influences resource densities and the resilience of food webs to perturbations.
To test the generality of these effects, I examined them under different assumptions of evolutionary tradeoffs, consumer foraging behavior, and resource distributions.
My general finding was that, whenever character displacement increased the strength of consumer-resource interactions, it resulted in lower resource densities and less resilient food webs. 
Character displacement always strengthened consumer-resource interactions whenever consumers were competing for resources that occured in distinct habitats. 
This occurred under different evolutionary tradeoffs and in both simple and more realistic foraging scenarios.
While character displacement did not always increase the strength of consumer-resource interactions, these contingent effects only occurred under the simplest and least realistic foraging scenario. 
<!--In the simple ecological scenario that has been the focus of past models of character displacement (type 1 functional response and competing for resources in the same habitat), I find that the shape of the tradeoff can qualitatively affect the relationship between character displacement, resource densities, and food-web stability. 
Extensions of this model, ones that include other foraging scenarios with more realistic functional responses (e.g. saturating) and that reflect the biological reality that consumers are often more mobile than their resources, I find predictable effects of character displacement on resource densities and food-web stability.
Competition theory has generally shown that character displacement enhances consumer coexistence.
I show that this adaptive process comes at the ecological cost of making food webs less resilient to perturbations.  -->
Taken together, my results indicate that the adaptive process of character displacement often comes at the ecological cost of making food webs less resilient to perturbations. 
They also suggest that food-web and competition theory can give complimentary insights to the ecological consequences of adaptation. <!-- eco-evolutionary feedbacks -->

@@

<!-- ECD is an important process, and current empirical and theoretical work suggests it often results in divergence of consumer foraging traits.-->

@@

<!-- Ecological consequences of character displacement have received surprisingly little attention. In particular, prior work doesn't allow a proper comparision of the effects of character displacement on community stability.-->

Although explicit models of resources gave insights to the evolution of character displacement, the ecological feedback onto consumer-resource dynamics have received surprisingly little attention. 
This is likely because the ecological feedback has been primarily studied through the lens of competition and coexistence theory [@Lawlor1976; @Germain2018; @Bassar2017; @McPeek2019]. 
For example, early theoretical work by @Lawlor1976 showed that ecological character displacement promotes coexistence by favoring specialized consumers that experience reduced interspecific competition. 
<!--
On the other hand, specialization often increases a consumer’s ability to capture a specific resource.
From food-web theory, we know that increased feeding rates often lead to resource suppression, which if sufficient enough, can generate oscillations, and hence less stable, consumer-resource dynamics [Rosenzweig 1971; Murdoch et al. 2003; McCann 2011].
-->
Yet, this reduction in interspecific competition may, at the same time, increase the interspecific interaction between a consumer and the resource it specializes on.  
Both food-web theory and empirical studies have shown that increasing the strength of consumer-resource interactions often leads to resource suppression, which if sufficient enough, can generate oscillations, and hence less stable consumer-resource dynamics [@Rosenzweig1971; @Luckinbill1973; @Murdoch2002; @Murdoch2003; @McCann2011].
Thus, a food-web perspective, which accounts for both the direct and indirect effects of consumer-resource interactions, may yield new insight to the ecological consequences of character displacement.

<!--Thus, a consumer-resource perspective may yield new insight to the ecological feedback of character displacement on food-web dynamics. 
<!--Given that character displacement is mediated by a consumer-resource interaction, a food-web perspective may yield   NEED A LAST SENTENCE TO WRAP IT UP, HIGHLIGHTING HOW A FOOD-WEB PERSPECTIVE MAY YIELD NEW ECOLOGICAL INSIGHTS AND GIVE A CONTRASTING PERSPECTIVE TO COEXISTENCE THEORY. PERHAPS A I NEED TO INCLUDE A SENTENCE MID-WAY THROUGH SUGGESTING THAT A FOOD-WEB PERSPECTIVE MAY GIVE QUALITATIVELY DIFFERENT INSIGHTS. IT MAY BE NICE TO BRING IN THE SPATIAL COUPLING WORK OF MCCANN SUGGESTING THAT SPECIALIZATION CAN COUPLE FOOD-WEBS, BUT THIS WAS ONLY WITH ONE TOP PREDATOR (I.E. NO CONSUMERS...LOOK AT OLD TEXT FOR WORDING SUGGESTIONS.)-->

<!-- older idea...Write 3rd paragraph of intro, highlighting how the direct effects of character displacement, surprisingly, have been ignored. We are currently lacking insight to the relative importance of these direct and indirect effects of character displacement on community dynamics.

<!--
The role of character displacement in governing food-web dynamics remains unclear A food-web perspective, however, has been lacking from theoretical work on the ecological effects of character displacement. , thus the consequences of this process for food-web stabilit

<!--
Despite the key insights made by explicit models of resource dynamics,

gave key insights to the evolutionary dynamics of character displacement, the ecological effects on resources dynamics have received surprisingly little attention. Perhaps this is because the ecological dynamics have been studied primarily through the lens of competition and coexistence thoery

This oversight is unfortunate, given that the stability of consumer-resources is often closely tied with resource abundances.

Although character displacement is the result of an eco-evolutionary feedback, the ecological consequences of this process have received surprisingly little attention [@Germain2018]. Perhaps this is because early theoretical work suggested that character displacement was a potent stabilizing force in ecological communities [@Lawlor1976]. This ecological outcome is intuitive in light of competition theory ---character divergence reduces interspecific competition between consumers, promoting coexistence. On the other hand, specialization often enhances a consumer’s ability to capture a specific resource. A well known consequence of increased foraging efficiency (all else equal) is resource suppression, which if sufficient enough, can generate oscillations, and hence less stable, consumer-resource dynamics (Rosenzweig 1971; Murdoch et al. 2003; McCann 2011).

Moreover, past inferences about the stabilizing effects of character displacement have only been made in communities where both consumers are present. However, I argue that inferences about the effect of character displacement, like the process itself, should only be made when comparing communities in the presence and absence of competition [criteria #3 for character displacement, @Schluter1992; @Schluter2000]. Therefore, the role of ecological character displacement in determining community stability remains unclear.

<!--
However, inferences about the effect of character displacement, like the process itself, can only be made when comparing communities in the presence and absence of competition [criteria #3 for character displacement, @Schluter1992; @Schluter2000]. Current inferences about the stabilizing effects of consumer divergence are only comparisons about communities where there are two consumers, and even though, the focus is only on the stable coexistence of consumer species.

<!--
The consequence of this eco-evolutionary feedback for consumer coexistence has been well characterized (Lawlor and Smith 1976; Germain's work). Ecological character displacement has typically been considered to be a stabilizing factor in communities (Lawlor and Smith 1976). This is because divergence in foraging traits is thought to reduce the strength of interspecific competition between consumers. Reducing the strength of interspecific competition relative to intraspecific competition is a stabilizing force in consumer-resource systems. Consumer specialization on distinct resources suggests that there are niche differences between consumers. Mathematically, this is also often assumed to be the case, because using techniques such as evolutionary invasion analysis describe the ability for a consumer with a mutant allele to increase in abundance when rare, a demographic signature of niche differentiation. Therefore, if both consumers are able to invade when rare, then this suggests that they can coexist.

Character displacement involves evolutionary shifts in character states in response to competition [@Schluter1992; @Schluter2000]. Theoretically, this is determined by comparing character changes in the presence and absence of a competitor. Empirically, this is determined by meeting several criteria that rule out other possible explanations [@Schluter1992; @Schluter2000]. Importantly, current inferences about the effect of character displacement on community stability do not meet this critera. For example, the inference that character displacement enhances species coexistence is only made in a scenario where consumers are competing. This type of stability is important, but it does not enable us to examine the effect of character displacement on community stability. As a consequence, I argue that the effect of character displacement on community stability remains an open question.

Of the prior models of character displacement, only @Lawlor1976 examine the effect of character displacement on the ecological stability of the system. They conducted a detailed theoretical analysis of three different metrics of stability: (i) stable coexistence of consumers, defined as the mutual invasibility of each consumer; (ii) global stability of consumers, defined as the continued coexistence of both consumers when resource dynamics are perturbed; and (iii) local stability, defined as the rate at which the four-species food web returned to equilibrium after a small perturbation to species abundances. Note that the first two metrics correspond to the stable coexistence of consumers, while the third metric measures the stability of the entire food web. @Lawlor1976 found that character displacement enhanced the stable coexistence of consumers, yet had a small negative effect on the local stability of the food web. 
@Lawlor1976's work gave fundamental insight to the effect of character displacement on ecological stability; however, none of their stability analyses compared food webs with and without a competing consumer. Such comparisons are necessary for inferring the effects of ecological character displacement, a point that has been made clear in the criteria to demonstrate character displacement (@Schluter2000; @Schluter1992). In fact, the first two metrics of stability can only be assessed when both consumers are present because it examines the effect of divergence on coexistence, which requires both consumers to be present. In contrast, local stability can be compared in food webs with and without a competing consumer, and thus provides a useful benchmark for examining the effects of character displacement on food-web stability.

<!-- -> Another paragraph illustrating how the direct and indirect effects of character displacement likely have constrasting effects on stability. -->

<!--
Indeed, ecological character displacement often results in the evolution of characters that enhance a consumer’s ability to capture resources, which tends to destabilize consumer-resource interactions (Murdoch et al. 2003; McCann 2011). This occurs because increases in a consumer’s attack rate results in the resource being suppressed well below its carrying capacity, which if sufficient enough, can generate oscillations, and hence less stable, consumer-resource dynamics (Rosenzweig 1971; Murdoch et al. 2003). Experimental decreases in interaction strength between consumers and resources has been shown to promote coexistence between consumers and resources (Luckinbill 1973). Moreover, there is broad support that specialized consumers, since they are tightly coupled to their resources, tend to exhibit consumer-resource cycles, whereas this is not the case for generalists (Murdoch et al. 2002). Therefore, although evolutionary responses to exploitative competition may reduce the strength of interactions between consumers, they ultimately increase the strength of interactions with their resources which can destabilize the direct consumer-resource interaction (rather than the indirect consumer-consumer interaction). While prior models of character displacement have included more realistic functional responses and spatial relationships between consumers and resources, they have focused more on whether these factors alter patterns of character displacement rather than the effects of character displacement on the stability of the ecological community. As a consequence, it is currently unclear whether ecological character displacement has a net stabilizing or destabilizing effect on ecological communities.

<!-- Here we fill this knowledge gap. We focus on the ecological side of these dynamics, because the evolutionary aspects have been well characterized, although we show in the appendices that the evolutionary dynamics are as expected (especially for McCann model). -->

@@

<!-- Explanation of model parameters -->

To examine how ecological character displacement affects resource abundances and food-web stability<!--consumer-resource interactions and food-web dynamics-->, I analyzed a continuous-time model of two consumers ($C_1,C_2$) competing for two shared resources ($R_1,R_2$):

@@

<!-- Methods for assessing resource abundances and stability. -->

@@

<!-- Explain analytical methods -->

I examine the effects of character displacement on food-web dynamics by (i) deriving analytical expressions for the relationship between consumer attack rates and food-web dynamics and (ii) simulating the effects of competition on the eco-evolutionary dynamics of consumers and resources. To gain analytical insight, I assume that resources are equivalent ($r=r_1=r_2$ and $K=K_1=K_2$) as well as consumers ($e=e_{11}=e_{12}=e_{21}=e_{22}$; $m=m_1=m_2$), except that consumer attack rates are mirror images of each other ($a_{ii}=a_{11}=a_{22}$; $a_{ij}=a_{12}=a_{21}$). While this scenario is arguably simplistic, it allows me to isolate the effects of ecological character displacement. <!--, rather than any dependence on inherent asymmetries between consumers or resources.-->

<!-- Explain simulations -->

@@

<!-- Redundant, removing for now
In what follows, I examine the ecological consequences of character displacement under three different attack-rate tradeoffs (n>1, n=1, n<1) and in three different foraging scenarios: (i) resources occur in the same habitat and consumers exhibit a linear functional response (e.g. @MacArthur1972); (ii) resources occur in different habitats and consumers exhibit a linear functional response (e.g. @Lawlor1976); (iii) resources occur in distinct habitats and consumers exhibit a more realistic functional response [@McCann2005]. I show that the consequences of character displacement for food-web dynamics are not intuitive and depend critically on the foraging scenario and tradeoff in consumer attack rates. 
-->

@@

<!-- Functional response equation and illustration of assumptions already given in previous section -->

<!-- Redundant, removing for now
The evolution of consumer attack rates in this model has been analyzed in detail [@Lawlor1976; @Abrams1986], with the general result being divergent character displacement. -->
When both consumers and resources are present, the density of resources at equilibrium are equivalent ($\hat R = \hat R_1 = \hat R_2$) and are determined by the following equation (derivation in supplementary material):

@@

The effect of character displacement on resource density closely matches its impact on food-web stability (Fig. \ref{fig:MacArthur_Plot}D). For example, when character displacement results in resource suppression (orange), there is a corresponding decrease in stability. Similarly, if character displacement does not influence resource densities (blue), then there is no corresponding effect on stability. Curiously, this match between impact on resources and stability breakdowns with the concave down tradeoff (green). <!--This is not simply a consequence of having an additional consumer in the system, but emerges from the eco-evolutionary feedback between character displacement and resource suppression (Fig. \ref{fig:Stability}). The effect of this eco-evolutionary feedback can be seen by comparing initial (small points) with evolved strategies (large points). For example, when the tradeoff is concave up (orange points), the three-species system (small triangle) is initially more stable than the four-species system (small circle), but this relationship reverses by the time consumers reach their evolutionary stable strategy.-->

@@

<!-- Functional response equation and illustration of assumptions -->

@@

<!-- Effect on attack rates, resource densities, and stability -->

@@

This equation implies that if consumers evolve to become specialists on resources that occur in their preferred habitat (e.g. $w>0.5$ and $a_{ii}>a_{ij}$), then the effective attack rate of consumers ($wa_{ii}+(1-w)a_{ij}$) will always increase, regardless of the tradeoff (Fig. \ref{fig:LS_Plot}A). Thus, character displacement always results in resource suppression if consumers are competing for resources that occur in different habitats (Fig. \ref{fig:LS_Plot}B). Note that the shape of the tradeoff can modify the effect of character displacement. This is not so much due to the tradeoff affecting the magnitude of displacement (it does, but the effect is minor), but because the form of the tradeoff affects resource abundances when a single consumer has reached its ESS (triangles in Fig. \ref{fig:LS_Plot}B). In contrast, resource densities reach a similar value when consumers evolve in the presence of a competitor (circles in Fig. \ref{fig:LS_Plot}B), because character displacement tends to reach a constraint of complete specialization. It is worth noting that in this foraging scenario, resource densities are consistently higher at the single consumer ESS compared to the predictions I derived for when both consumers are present (deviation of triangles from respective colored lines in Fig. \ref{fig:LS_Plot}B). This is likely because consumers actually evolve to be slightly specialized on the resources that occur in their non-preferred habitat (deviation of triangles from 0.5 along x-axis in Fig. \ref{fig:LS_Plot}B). <!-- I will revisit this subtle point when I examine the effects of character displacement on food-web stability. -->

@@

```{r LS_Resources, include=FALSE, fig.cap="\\label{fig:LS_Resources}Effect of character displacement on effective attack rates (A) and resource densities (B) when consumers cannot forage for resources simultaneously."}
LS_attack <- ggplot(tradeoff_df, aes(x=aii/(aii+aij), y=wii*aii+wij*aij, color = factor(n))) + 
  geom_line() +
  geom_point(data = filter(LawlorSmith_displacements, Time=="End"), aes(x = a11/(a11+a12), y=w11*a11+w12*a12, fill=factor(n), shape=Competitor), size=3, color="white") + #, show.legend=FALSE) +
  geom_point(data = filter(LawlorSmith_displacements, Time=="Begin"), aes(x = a11/(a11+a12), y=w11*a11+w12*a12, fill=factor(n), shape=Competitor), size=1.5, color="white") +
  ylab(expression(w*a[ii]+(1-w)*a[ij])) +
  xlab(expression(frac(a[ii],a[ii]+a[ij]))) +
  scale_shape_manual(values=c(24,21)) +
  scale_color_manual(name="Tradeoff (n)", values=cbPalette)  +
  scale_fill_manual(name="Tradeoff (n)", values=cbPalette) +
  guides(fill = guide_legend(override.aes=list(shape=21)))

LS_resources <- cbind.data.frame(select(tradeoff_df, n, aii, aij), LawlorSmith_2C_2R_tradeoff_dynamics) %>%
  ggplot(., aes(x=aii/(aii+aij), y=R1+R2, color=factor(n))) + 
  geom_line() +
  geom_point(data = filter(LawlorSmith_displacements, Time=="End"), aes(x = a11/(a11+a12), y=R1+R2, fill=factor(n), shape=Competitor), size=3, color="white") + #, show.legend=FALSE) +
  geom_point(data = filter(LawlorSmith_displacements, Time=="Begin"), aes(x = a11/(a11+a12), y=R1+R2, fill=factor(n), shape=Competitor), size=1.5, color="white") +
  ylab(expression(R[i]+R[j])) +
  xlab(expression(frac(a[ii],a[ii]+a[ij]))) +
  scale_shape_manual(values=c(24,21)) +
  scale_color_manual(name="Tradeoff (n)", values=cbPalette)  +
  scale_fill_manual(name="Tradeoff (n)", values=cbPalette) +
  guides(fill = guide_legend(override.aes=list(shape=21)))

@@

```{r LS Model, include=FALSE}
plot_LS_model <- plot_grid(
  #tradeoff_plot + theme(legend.position = "none"), 
  LS_attack  + theme(legend.position = "none"), 
  LS_resources  + theme(legend.position = "none"),
  Stability_LS  + theme(legend.position = "none"), ncol=2, align='hv', labels = "AUTO")

@@

In the previous models, character displacement influences food-web stability, but all of the food webs ultimately return to a stable equlibrium (because $\lambda_{max}<0$). In this more realistic model, however, whether the food web is locally stable depends on the parameters of both consumers and resources. Specifically, <!--I conducted the previous simulations on a scenario where the food web remains locally stable ($\operatorname{Re}(\lambda_{max})<0$). Is it possible that character displacement would ever result in a locally unstable food web? To gain analytical insight to this question, I examined the conditions governing the local stability of the four-species food web when consumers exhibit a more realistic functional response (@McCann2005). I show that -->the four-species food web will transition from having a locally stable equilibrium to a limit cycle under the following conditions (derivation in supplementary material):

@@

Previously, I showed that character displacement always increases the effective attack rate of consumers ($wa_{ii}+(1-w)a_{ij}$), regardless of the the shape of the tradeoff (Fig. \ref{fig:LS_Plot}A). Thus, character displacement always pushes the food web toward the boundary of local stability under this more realistic foraging scenario. <!--Whether or not the food web becomes locally unstable, however, depends on the choice of initial parameters.-->

```{r McCann_Resources, include=FALSE, fig.cap="\\label{fig:McCann_Resources}Effect of character displacement on total attack rates (A) and resource densities (B) when consumers exhibit a more realistic functional response."}
McCann_attack <- ggplot(tradeoff_df, aes(x=aii/(aii+aij), y=wii*aii+wij*aij, color = factor(n))) + 
  geom_line() +
  geom_point(data = filter(McCann_displacements, Time=="End"), aes(x = a11/(a11+a12), y=w11*a11+w12*a12, fill=factor(n), shape=Competitor), size=3, color="white") + #, show.legend=FALSE) +
  geom_point(data = filter(McCann_displacements, Time=="Begin"), aes(x = a11/(a11+a12), y=w11*a11+w12*a12, fill=factor(n), shape=Competitor), size=1.5, color="white") +
  ylab(expression(w*a[ii]+(1-w)*a[ij])) +
  xlab(expression(frac(a[ii],a[ii]+a[ij]))) +
  scale_shape_manual(values=c(24,21))  +
  scale_color_manual(name="Tradeoff (n)", values=cbPalette)  +
  scale_fill_manual(name="Tradeoff (n)", values=cbPalette) +
  guides(fill = guide_legend(override.aes=list(shape=21)))

McCann_resources <- cbind.data.frame(select(tradeoff_df, n, aii, aij), McCann_2C_2R_tradeoff_dynamics) %>%
  ggplot(., aes(x=aii/(aii+aij), y=R1+R2, color = factor(n))) + 
  geom_line() +
  geom_point(data = filter(McCann_displacements, Time=="End"), aes(x = a11/(a11+a12), y=R1+R2, fill=factor(n), shape=Competitor), size=3, color="white") + #, show.legend=FALSE) +
  geom_point(data = filter(McCann_displacements, Time=="Begin"), aes(x = a11/(a11+a12), y=R1+R2, fill=factor(n), shape=Competitor), size=1.5, color="white") +
  ylab(expression(R[i]+R[j])) +
  xlab(expression(frac(a[ii],a[ii]+a[ij]))) +
  scale_shape_manual(values=c(24,21))  +
  scale_color_manual(name="Tradeoff (n)", values=cbPalette)  +
  scale_fill_manual(name="Tradeoff (n)", values=cbPalette) +
  guides(fill = guide_legend(override.aes=list(shape=21)))

@@

```{r Stability_McCann, include=FALSE, fig.cap="\\label{fig:Stability_McCann}Effect of character displacement on local stability when consumers exhibit a more realistic functional response."}
Stability_McCann <- cbind.data.frame(select(tradeoff_df, n, aii, aij), McCann_2C_2R_tradeoff_dynamics) %>%
  ggplot(., aes(x=aii/(aii+aij), y=-1*max.Re.eigen, color=factor(n))) + 
  geom_line() +
  geom_point(data = filter(McCann_displacements, Time=="End"), aes(x = a11/(a11+a12), y=-1*max.Re.eigen, fill=factor(n), shape=Competitor), size=3, color="white", show.legend = FALSE) +
  geom_point(data = filter(McCann_displacements, Time=="Begin"), aes(x = a11/(a11+a12), y=-1*max.Re.eigen, fill=factor(n), shape=Competitor), size=1.5, color="white") +
  xlab(expression(frac(a[ii],a[ii]+a[ij]))) +
  ylab(expression("Stability "(-lambda[max]))) + #ylab(expression(-1%*%"Re"(lambda[max]))) +
  scale_shape_manual(values=c(24,21))  +
  scale_color_manual(name="Tradeoff (n)", values=cbPalette)  +
  scale_fill_manual(name="Tradeoff (n)", values=cbPalette) +
  geom_hline(yintercept=0, linetype="dotted", color="grey") +
  guides(fill = guide_legend(override.aes=list(shape=21)))

>>>>

@@

Ecological character displacement is an adaptive process, often enhancing coexistence between consumers by reducing competition for shared resources. 
This focus on competition has undermined the fact that consumer-resource interactions ultimately mediate the process of character displacement. 
As a consequence, the effects of character displacement on consumer-resource interactions and food-web dynamics have not been critically examined.
Here, I study a model of two consumers competing for two shared resources to examine how character displacement in consumers influences resource densities and the resilience of food webs to perturbations.
To test the generality of these effects, I examined them under different assumptions of evolutionary tradeoffs, consumer foraging behavior, and resource distributions.
My general finding was that, whenever character displacement increased the strength of consumer-resource interactions, it resulted in lower resource densities and less resilient food webs. 
Character displacement always strengthened consumer-resource interactions whenever consumers were competing for resources that occured in distinct habitats. 
This occurred under different evolutionary tradeoffs and in both simple and more realistic foraging scenarios.
While character displacement did not always increase the strength of consumer-resource interactions, these contingent effects only occurred under the simplest and least realistic foraging scenario. 
Taken together, my results indicate that the adaptive process of character displacement often comes at the ecological cost of making food webs less resilient to perturbations. 
They also suggest that food-web and competition theory can give complimentary insights to the ecological consequences of adaptation.

@@

@@

Although explicit models of resources gave insights to the evolution of character displacement, the ecological feedback onto consumer-resource dynamics have received surprisingly little attention. 
This is likely because the ecological feedback has been primarily studied through the lens of competition and coexistence theory [@Lawlor1976; @Germain2018; @Bassar2017; @McPeek2019]. 
For example, early theoretical work by @Lawlor1976 showed that ecological character displacement promotes coexistence by favoring specialized consumers that experience reduced interspecific competition. 
Yet, this reduction in interspecific competition may, at the same time, increase the interspecific interaction between a consumer and the resource it specializes on.  
Both food-web theory and empirical studies have shown that increasing the strength of consumer-resource interactions often leads to resource suppression, which if sufficient enough, can generate oscillations, and hence less stable consumer-resource dynamics [@Rosenzweig1971; @Luckinbill1973; @Murdoch2002; @Murdoch2003; @McCann2011].
Thus, a food-web perspective, which accounts for both the direct and indirect effects of consumer-resource interactions, may yield new insight to the ecological consequences of character displacement.

@@

To examine how ecological character displacement affects resource abundances and food-web stability, I analyzed a continuous-time model of two consumers ($C_1,C_2$) competing for two shared resources ($R_1,R_2$):

@@

@@

I examine the effects of character displacement on food-web dynamics by (i) deriving analytical expressions for the relationship between consumer attack rates and food-web dynamics and (ii) simulating the effects of competition on the eco-evolutionary dynamics of consumers and resources. To gain analytical insight, I assume that resources are equivalent ($r=r_1=r_2$ and $K=K_1=K_2$) as well as consumers ($e=e_{11}=e_{12}=e_{21}=e_{22}$; $m=m_1=m_2$), except that consumer attack rates are mirror images of each other ($a_{ii}=a_{11}=a_{22}$; $a_{ij}=a_{12}=a_{21}$). While this scenario is arguably simplistic, it allows me to isolate the effects of ecological character displacement.

@@

@@

When both consumers and resources are present, the density of resources at equilibrium are equivalent ($\hat R = \hat R_1 = \hat R_2$) and are determined by the following equation (derivation in supplementary material):

@@

The effect of character displacement on resource density closely matches its impact on food-web stability (Fig. \ref{fig:MacArthur_Plot}D). For example, when character displacement results in resource suppression (orange), there is a corresponding decrease in stability. Similarly, if character displacement does not influence resource densities (blue), then there is no corresponding effect on stability. Curiously, this match between impact on resources and stability breakdowns with the concave down tradeoff (green).

@@

@@

@@

This equation implies that if consumers evolve to become specialists on resources that occur in their preferred habitat (e.g. $w>0.5$ and $a_{ii}>a_{ij}$), then the effective attack rate of consumers ($wa_{ii}+(1-w)a_{ij}$) will always increase, regardless of the tradeoff (Fig. \ref{fig:LS_Plot}A). Thus, character displacement always results in resource suppression if consumers are competing for resources that occur in different habitats (Fig. \ref{fig:LS_Plot}B). Note that the shape of the tradeoff can modify the effect of character displacement. This is not so much due to the tradeoff affecting the magnitude of displacement (it does, but the effect is minor), but because the form of the tradeoff affects resource abundances when a single consumer has reached its ESS (triangles in Fig. \ref{fig:LS_Plot}B). In contrast, resource densities reach a similar value when consumers evolve in the presence of a competitor (circles in Fig. \ref{fig:LS_Plot}B), because character displacement tends to reach a constraint of complete specialization. It is worth noting that in this foraging scenario, resource densities are consistently higher at the single consumer ESS compared to the predictions I derived for when both consumers are present (deviation of triangles from respective colored lines in Fig. \ref{fig:LS_Plot}B). This is likely because consumers actually evolve to be slightly specialized on the resources that occur in their non-preferred habitat (deviation of triangles from 0.5 along x-axis in Fig. \ref{fig:LS_Plot}B).

@@

```{r LS_Resources, include=FALSE, fig.cap="\\label{fig:LS_Resources}Effect of character displacement on effective attack rates (A) and resource densities (B) when consumers cannot forage for resources simultaneously."}
LS_attack <- ggplot(tradeoff_df, aes(x=aii/(aii+aij), y=wii*aii+wij*aij, color = factor(n))) + 
  geom_line() +
  geom_point(data = filter(LawlorSmith_displacements, Time=="End"), aes(x = a11/(a11+a12), y=w11*a11+w12*a12, fill=factor(n), shape=Competitor), size=3, color="white") + 
  geom_point(data = filter(LawlorSmith_displacements, Time=="Begin"), aes(x = a11/(a11+a12), y=w11*a11+w12*a12, fill=factor(n), shape=Competitor), size=1.5, color="white") +
  ylab(expression(w*a[ii]+(1-w)*a[ij])) +
  xlab(expression(frac(a[ii],a[ii]+a[ij]))) +
  scale_shape_manual(values=c(24,21)) +
  scale_color_manual(name="Tradeoff (n)", values=cbPalette)  +
  scale_fill_manual(name="Tradeoff (n)", values=cbPalette) +
  guides(fill = guide_legend(override.aes=list(shape=21)))

LS_resources <- cbind.data.frame(select(tradeoff_df, n, aii, aij), LawlorSmith_2C_2R_tradeoff_dynamics) %>%
  ggplot(., aes(x=aii/(aii+aij), y=R1+R2, color=factor(n))) + 
  geom_line() +
  geom_point(data = filter(LawlorSmith_displacements, Time=="End"), aes(x = a11/(a11+a12), y=R1+R2, fill=factor(n), shape=Competitor), size=3, color="white") + 
  geom_point(data = filter(LawlorSmith_displacements, Time=="Begin"), aes(x = a11/(a11+a12), y=R1+R2, fill=factor(n), shape=Competitor), size=1.5, color="white") +
  ylab(expression(R[i]+R[j])) +
  xlab(expression(frac(a[ii],a[ii]+a[ij]))) +
  scale_shape_manual(values=c(24,21)) +
  scale_color_manual(name="Tradeoff (n)", values=cbPalette)  +
  scale_fill_manual(name="Tradeoff (n)", values=cbPalette) +
  guides(fill = guide_legend(override.aes=list(shape=21)))

@@

```{r LS Model, include=FALSE}
plot_LS_model <- plot_grid(
  LS_attack  + theme(legend.position = "none"), 
  LS_resources  + theme(legend.position = "none"),
  Stability_LS  + theme(legend.position = "none"), ncol=2, align='hv', labels = "AUTO")

@@

In the previous models, character displacement influences food-web stability, but all of the food webs ultimately return to a stable equlibrium (because $\lambda_{max}<0$). In this more realistic model, however, whether the food web is locally stable depends on the parameters of both consumers and resources. Specifically, the four-species food web will transition from having a locally stable equilibrium to a limit cycle under the following conditions (derivation in supplementary material):

@@

Previously, I showed that character displacement always increases the effective attack rate of consumers ($wa_{ii}+(1-w)a_{ij}$), regardless of the the shape of the tradeoff (Fig. \ref{fig:LS_Plot}A). Thus, character displacement always pushes the food web toward the boundary of local stability under this more realistic foraging scenario.

```{r McCann_Resources, include=FALSE, fig.cap="\\label{fig:McCann_Resources}Effect of character displacement on total attack rates (A) and resource densities (B) when consumers exhibit a more realistic functional response."}
McCann_attack <- ggplot(tradeoff_df, aes(x=aii/(aii+aij), y=wii*aii+wij*aij, color = factor(n))) + 
  geom_line() +
  geom_point(data = filter(McCann_displacements, Time=="End"), aes(x = a11/(a11+a12), y=w11*a11+w12*a12, fill=factor(n), shape=Competitor), size=3, color="white") + 
  geom_point(data = filter(McCann_displacements, Time=="Begin"), aes(x = a11/(a11+a12), y=w11*a11+w12*a12, fill=factor(n), shape=Competitor), size=1.5, color="white") +
  ylab(expression(w*a[ii]+(1-w)*a[ij])) +
  xlab(expression(frac(a[ii],a[ii]+a[ij]))) +
  scale_shape_manual(values=c(24,21))  +
  scale_color_manual(name="Tradeoff (n)", values=cbPalette)  +
  scale_fill_manual(name="Tradeoff (n)", values=cbPalette) +
  guides(fill = guide_legend(override.aes=list(shape=21)))

McCann_resources <- cbind.data.frame(select(tradeoff_df, n, aii, aij), McCann_2C_2R_tradeoff_dynamics) %>%
  ggplot(., aes(x=aii/(aii+aij), y=R1+R2, color = factor(n))) + 
  geom_line() +
  geom_point(data = filter(McCann_displacements, Time=="End"), aes(x = a11/(a11+a12), y=R1+R2, fill=factor(n), shape=Competitor), size=3, color="white") + 
  geom_point(data = filter(McCann_displacements, Time=="Begin"), aes(x = a11/(a11+a12), y=R1+R2, fill=factor(n), shape=Competitor), size=1.5, color="white") +
  ylab(expression(R[i]+R[j])) +
  xlab(expression(frac(a[ii],a[ii]+a[ij]))) +
  scale_shape_manual(values=c(24,21))  +
  scale_color_manual(name="Tradeoff (n)", values=cbPalette)  +
  scale_fill_manual(name="Tradeoff (n)", values=cbPalette) +
  guides(fill = guide_legend(override.aes=list(shape=21)))

@@

```{r Stability_McCann, include=FALSE, fig.cap="\\label{fig:Stability_McCann}Effect of character displacement on local stability when consumers exhibit a more realistic functional response."}
Stability_McCann <- cbind.data.frame(select(tradeoff_df, n, aii, aij), McCann_2C_2R_tradeoff_dynamics) %>%
  ggplot(., aes(x=aii/(aii+aij), y=-1*max.Re.eigen, color=factor(n))) + 
  geom_line() +
  geom_point(data = filter(McCann_displacements, Time=="End"), aes(x = a11/(a11+a12), y=-1*max.Re.eigen, fill=factor(n), shape=Competitor), size=3, color="white", show.legend = FALSE) +
  geom_point(data = filter(McCann_displacements, Time=="Begin"), aes(x = a11/(a11+a12), y=-1*max.Re.eigen, fill=factor(n), shape=Competitor), size=1.5, color="white") +
  xlab(expression(frac(a[ii],a[ii]+a[ij]))) +
  ylab(expression("Stability "(-lambda[max]))) + 
  scale_shape_manual(values=c(24,21))  +
  scale_color_manual(name="Tradeoff (n)", values=cbPalette)  +
  scale_fill_manual(name="Tradeoff (n)", values=cbPalette) +
  geom_hline(yintercept=0, linetype="dotted", color="grey") +
  guides(fill = guide_legend(override.aes=list(shape=21)))
